---
title: "Learn, Visualize, and Analyze!"
output: html_document
date: "2025-05-13"
author: "Laforest-Lapointe Lab"
---

### Load Packages And Data
This is a test to practice pushing modifications on scripts with GitHub.
Test2

#### Load packages   
Run the `install.packages()` if packages are not installed yet.
```{r results='hide', message=FALSE}
#install.packages("ggpubr")
library(ggpubr)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("FSA")
library(FSA)
#install.packages("dunn.test")
library(dunn.test)
#install.packages("dplyr")
library(dplyr)
#install.packages("rstatix")
library(rstatix)
```

```{r include=FALSE}
assay <- read.table("/Users/bous2941/Library/CloudStorage/OneDrive-USherbrooke/PhD_SOPHIE_BOUTIN/TEACHING/RWORKSHOP_2025/assay.txt")
```

#### Import data
```{r eval=FALSE}
assay <- read.table("assay.txt")
```
Change variables class to factor.
```{r}
assay$Antibiotic <- as.factor(assay$Antibiotic)
assay$OD <- as.factor(assay$OD)
```


## Compare the mean of two groups

**Conditions to apply t-test (parametric)** 

1. Quantitative variable 

2. Normality 

To test normality: Shapiro-Wilk `shapiro.test()` 

3. Equality of variances 

To test equality of variance: Fisher-Snedecor (F test) `var.test()`

4. Independent observations 

**What to do if conditions are not respected:**

- Not normal (big *n*) 

*t*-test is fine but its power is reduced 

We can normalise the data 

permutation *t*-test is better 

- Not normal (small *n*) 

Permutation *t*-test if variances are equal, if not -> Wilcoxon-Mann-Whitney (*U*) 

- Semi-quantitative 

Wilcoxon-Mann-Whitney (*U*)


## Compare the mean across more than two groups

### ANOVA (parametric)

**Conditions to apply t-test (parametric)** 

1. Independent observation  

2. Normality (`shapiro.test()`) 

3. Homogeneous variance (`leveneTest()`) 

#### post-hoc HSD Tukey

-> Same conditions as ANOVA  


### Kruskal-Wallis (non-parametric)

#### Dunn test






## Correlation between two variables

Pearson
Spearman
Permutation









### T-test

#### Make a boxplot of CFU in function of antibiotic type
```{r}
(plot1 <- ggplot(assay, aes(Antibiotic, CFU, colour = Antibiotic, fill = Antibiotic))+
  geom_boxplot()+
  theme_classic()+
  scale_color_manual(values = c("black", "black"))+
  scale_fill_manual(values = c("lightskyblue","palegreen3")))
```

#### Perform t-test
```{r}
t.test(assay$CFU~assay$Antibiotic) # p-value = 0.06063
```
Create an object with the t-test output
```{r}
res.t.test <- t.test(assay$CFU~assay$Antibiotic)
```
Extract p-value from t-test output
```{r}
res.t.test$p.value
```

#### Different ways to add t-test p-value on graph
```{r}
plot1 +
  # Using stat_compare_means
  stat_compare_means(method = "t.test")+
  # Using annotate and the t-test output
  annotate("text", x = 0.6, y =30, label = res.t.test$p.value, color = "red")+
  # Same but round it to 2 digits
  annotate("text", x = 0.6, y =27, label = round(res.t.test$p.value, digits = 2), color = "blue")+
  # Add value manually (Not "automatic")
  annotate("text", x = 1.5, y =30, label = "T-test, p = 0.06", color = "darkgreen")
```

### Anova, Kruskal, Dunn

#### Make a plot of CFU count in function of optical density (OD)
```{r}
(plot2 <- ggplot(assay, aes(OD, CFU, color = OD, fill = OD))+
  stat_summary(geom = "col", fun = mean)+
  theme_classic()+
  ggtitle("Bacterial Count")+
  scale_color_manual(values = c("black", "black", "black"))+
  scale_fill_manual(values = c("darkolivegreen1","darkolivegreen3", "darkolivegreen4")))
```

Add points
```{r}
(plot2 +
  geom_jitter(show.legend = FALSE))
```

Add error bars
```{r}
(plot2 <- plot2 +
  geom_jitter(show.legend = FALSE)+
  stat_summary(geom = "errorbar", fun.data = "mean_sdl", fun.args = list(mult = 1), width = 0.5, show.legend = FALSE))
```

#### Perform statistical tests
```{r}
res.aov <- aov(CFU ~ OD, data = assay)
summary(res.aov) # p = 9.5x10(-16)
kruskal.test(CFU ~ OD, data = assay) # p = 1.5x10(-9)
dunn_results <- dunnTest(CFU ~ OD, data = assay, method = "bonferroni")
```
Add a column showing significance levels (***) to the Dunn test result object
```{r}
dunn_results$res
dunn_results$res <- dunn_results$res %>%
  mutate(stars = case_when(
    P.adj < 0.001 ~ "***",
    P.adj < 0.01 ~ "**",
    P.adj < 0.05 ~ "*",
    TRUE ~ "ns"
  ))
```

#### Add stats on plots
Some parameters needed for the next plots
```{r}
compar <- lapply(strsplit(dunn_results$res$Comparison, " - "), function(x) x)
y_positions <- c(max(assay$CFU) + 0.5, max(assay$CFU) + 3.5, max(assay$CFU) + 7)
```

Use *stat_compare_means( )*
```{r}
plot2+
  stat_compare_means(method = "kruskal.test")

plot2+
  stat_compare_means(method = "anova")
```

Use *geom_signif( )*
```{r}
plot2+
  geom_signif(comparisons = compar, 
              annotations = format(dunn_results$res$P.adj, scientific = TRUE),
              textsize = 4,  
              y_position = y_positions) 

plot2 +
  geom_signif(comparisons = compar, 
              annotations = dunn_results$res$stars,
              textsize = 5,  
              y_position = y_positions)
```


```{r}
```









