---
title: "Introduction to R"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: true
    toc_depth: 3
    number_sections: true
    theme: cosmo
    highlight: tango
author: "Isabelle Laforest-Lapointe, Amy Heim, Sophie Boutin, Jonathan Rondeau-Leclaire"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This tutorial was created for a 3-day free R introduction workshop organized by the [Laforest-Lapointe Lab](https://www.laforestlab.com/) at Université de Sherbrooke on May 12th-14th.

# Introduction

R is an object-oriented programming language. Essentially, this means that data, values, inputs and outputs are saved as "objects" whose names are defined by the user. Objects can then be used to perform calculations, data manipulation, or visualisations.

## RStudio Environment

RStudio is a program that allows you to write a series of command as a script, have R execute these commands, visualise the output and see the objects that you create.

When you open RStudio for the first time, you will have three panels.

**Top left: Script**

Document where you will write the command lines. To create a script: `File > New File > R Script`. Save regularly!

Execute commands, place your cursor in the command's line and press: `Command + Enter` for Mac, `Control + Enter` for Windows.

**Bottom left: Console **

Where the output of the command line will appear. When you execute commands from the script, the command and its output (or error messages!) will appear there.

Use it to execute commands that you do not want to save in the script or to test your commands before adding them to the script. To execute a command from the console, simply press `enter` after writing it.

**Top right: Environment**

Place where all the objects you create (dataframes, values, vectors) are shown. Click on any object to visualise it.

**Bottom right: Files, Plots, Help**

This is a files and plots viewer as well as the help center.

## Basic R usage

Use R script as a calculator and get the output from the console:
```{r}
# Everything written after a hashtag is a comment (cannot be executed)
2+4
```

Store the output in an object
```{r}
res <- 2+4
res
```

Multiply the result
```{r}
2*res
```

## Syntax
R has a bunch of symbols with specific functions. Some are absolutely essential to know, which we present here.

### Common symbols

- `$` access an element of an object by its name (e.g. the column of a data frame)  
- `[]` access an element of an object by its index (e.g. `vector[3]` accesses the 3rd element in than vector)  
- `<-` assign an object (shortcut: `alt` + `-` on Windows, `option`+ `-` on mac)  
- `=` assign an object; specify function parameters  
- `""` and `''` are used to express strings (character values) 
- `?` used to learn how a function work. `?sum` will open a help file where you can learn about how to use that function.

### Comparison operators
To compare values to one another:

- `==` equal  
- `!=` is not equal  

The following can only be used for numeric values:

- `<` smaller than  
- `>` greater than    
- `<=` smaller or equal to  
- `>=` greater or equal to  

```{r}
5 > 6
5 <= 6
```

``TRUE`` and ``FALSE`` are boolean values with a specific meaning. They represent a type of information that can only be true or false. 

```{r}
5 != 4+1
```

### Logical operators
  
- `&` and  
- `|` or

They are used to test multiple conditions at once
```{r}
(5 == 5) & (6 == 6)
(5 == 4) & (6 == 6)
(5 == 4) | (6 == 6)
```

## Object classes

Objects and object components (e.g., data frame variables) always have a class. This class gives you a hint on how it will behave.

There exist many object classes, but a few are much more common than others. In the above example with `str()`, we can see two other classes, namely `num` for numeric vectors and `factor`, which is a special (composite) class for categorical data. Other common classes are `character` (chr), `integer` (int), `character` (chr) and `logical` (logi).

We can also see a composite class, `data.frame`. These are used for more complex objects, usually with two dimensions. Other composite classes include `matrix`, which are similar to data frames (the difference between these two is that data frames can contain elements of various classes). There are also `list` and `array`, which won't be covered in this course.

`Character`  
`Numeric`  
`Integer`  
`Factor`  
`Logical`  
`Date`  
`Vector`  
`Data frame`  
`Matrix`  

Let's create a few objects and look at the class R will assign them by default. There are a few basic functions worth learning about at this point. Functions are little programs that do specific things and you'll be using them all the time. The most basic function is `c()` which is used to combine its arguments into a vector. If all arguments are numeric, the vector will be numeric. If they contain at least one string element (wrapped in `"quotes"`), it will be a character vector. 

```{r}
# Create a vector of numeric values
vec1 <- c(1,3,5)

# Create a vector of characters values
vec2 <- c("one","three","five")

# Create a vector of logical values
vec3 <- c(TRUE, TRUE, FALSE)

# Bind these vectors into a dataframe
df1 <- data.frame(vec1, vec2, vec3)
class(vec1) # Use function class( ) to get the class of an object
class(vec2)

# Change the class of vec2 to factor
vec2 <- as.factor(vec2) 
vec2
class(vec2)
class(df1)

# Look at the structure of the data frame we made:
str(df1)
```

## Save an object

```
# Save object
saveRDS(df1, "df1.RDS")
# Read/open object
df1_new <- readRDS("df1.RDS")
```

## Challenge N˚1  

Create the following objects with 4 made-up values each:  

- `sample_id`: a character vector 
- `CFU`: a numeric vector  
- `Control`: a boolean vector telling us which samples are controls  
- `Date` : a vector of dates with format "YYYY-MM-DD"  

Then,  

1. combine these into a data frame
2. using `str()`, check the class of your data frame columns
4. Save your dataframe as an RDS file

# Getting started with data

Now let's import a small dataset and familiarise ourselves with a few functions that will help us look at our data in more detail.

## Working directory

It's good to have a working directory specifically dedicated to your project, where you store your data, scripts, output, etc. This way you can organise what goes in and comes out of R and keep track of your various scripts.

R can look anywhere in our computer for files. However, to load a file, you need to know its full path, e.g. `'/Users/jorondo/Repos/RworkshopUdS_2025/data.txt'` which can be annoying to write out. Instead, it's useful to set your working directory once at the beginning of your script, and then use relative paths to find your file. To do this, start by creating a folder in your computer, from where to work:

```
dir.create('~/Rworkshop2025') # Note that paths are always quoted  
setwd('~/RWorkshop2025')
dir.create('Output')
```
The `~` is a shortcut to your Home folder. You can always create your folders by hand, but it helps to have the path in your script. Now, if we save an object we previously created,

```
saveRDS(my_first_data_frame, "Output/my_first_data_frame.RDS")
```

It will be saved in that directory without needing to specify the absolute path.

## Data inspection

In any project, a first step should always be to look at your data as it appears in R. Many functions exist to help you do that. `summary()` will summarise variables in a class-dependent manner: numeric variables get a numeric summary (median, range, etc.), factors and logical get a count per category, and characters get a overview of the first values.

```{r, warning = FALSE, message = FALSE}
assay <- read.table("data/assay.txt", header = TRUE)
summary(assay)
```

Other useful functions are `head()`, which shows the first rows of an object, and `View()`, which opens shows the object in a formatted manner, a bit like an excel sheet.

```{r head}
head(assay)
```

Other useful checks are `dim()` or `length()`, especially when playing with functions that modify our objects:

```{r dim}
dim(assay) # shows the dimensions of matrices and data frames
ncol(assay) # number of columns
colnames(assay) # column names (our variable names!)
length(assay$OD) # check the length of a vector (a column fetched using $ is a vector)
```

## Basic operations

R can be used to operate on values in pretty much any way imaginable, from very simple things like calculating the mean to more complex things like removing certain characters from every values in a data frame variable.

Most mathematical operations are available as functions :

```{r math ops}
some_numbers <- head(assay$CFU)
some_numbers
sum(some_numbers) # Sum the values of a numeric vector
sqrt(some_numbers) # square root of each value
sum(some_numbers)/length(some_numbers) # compute the mean "manually"
mean(some_numbers) # compute the mean using a specific function
median(some_numbers)
sd(some_numbers) # compute the standard deviation
min(some_numbers) # find the minimum
```

Operations can also be done between columns of a data frame. In the following, we multiply two columns together and add the resulting vector as a new column on the same dataset using the `<-` (assignment) operator.

```{r}
assay$new_column <- assay$CFU * assay$OD # with $ sign
assay['new_column'] <- assay$CFU * assay$OD # equivalent!
summary(assay) 
```

### Loops

Loops are very useful, as they help us do similar operations multiple times while writing them only once. Let's say there you would like to multiply three columns of your dataframe by 1000. If your data was a numeric matrix, you could simply multiply the matrix object by 1000, e.g. `new_matrix = old_matrix*1000`. But we usually work with data frames, because our data contains various types of variables. Thus, columns may be characters or factors, which obviously cannot be multiplied. So we want to selectively multiply certain columns :

```{r}
my_data_frame <- assay # we just copy the df not to overwrite it
my_data_frame$CFU <- my_data_frame$CFU*1000
my_data_frame$OD <- my_data_frame$OD*1000
my_data_frame$new_column <- my_data_frame$new_column*1000
```

and so on for every variable. Using a loop, it would look like this:

```{r}
for (i in c('OD', 'CFU', 'new_column') ) { # define what to iterate on
  my_data_frame[[i]] <- my_data_frame[[i]]*1000
}
summary(my_data_frame)
```

It may not be useful for such a simple operation on only a few columns, but you can imagine how much time would be saved if the operations you do are more complex and need to be applied to multiple columns. Moreover, if you realise you need to multiply by 100 instead, you only have to change in one place, reducing the risk of syntax error.

## Challenge N˚2



## R as a calculator

You can also straight up use R as you would a calculator. Let's find the area of a single slice of a 14" pizza cut in 8, then format the result to be more "readable" :

```{r}
slice_area <- (pi*(14/2)^2)/4
slice_area_formatted <- round(slice_area, 2)
paste('The slice of a 14 inches pizza cut in 8 will have a surface of', slice_area_formatted, "square inches.")
```

Notice that we've created many objects to do this. We could just as well have done a single-liner that only prints the result to our console:

```{r}
paste('The slice of a 14 inches pizza cut in 8 will have a surface of', round((pi*(14/2)^2)/4, 2), "square inches.")
```

but it's up to you to decide how explicit you want your code to be. More explicit code is more readable, but requires more steps to be written out. It's all about balance.

You can also use R as a random number generator. The simplest way is using the uniform distribution, `runif()`, which draws `n` observations between specified values, and all values have an equal chance of being drawn.

```{r}
a_vector_of_values <- runif(n = 10, min = 0, max = 100)
a_vector_of_values
sort(a_vector_of_values)
```

You can draw from the normal distribution too, by specifying the mean and sd :

```{r}
a_normal_draw <- rnorm(n = 100, mean = 0, sd = 1)
head(a_normal_draw)
hist(a_normal_draw)
```

# Do's and don'ts

## Importing new data

Often we save our data in excel files. We can install a package to get a new function specialised in reading excel files by running `install.packages('readxl')`. Once you've done this, add the following line at the top of your script:

```{r}
library(readxl) # you need execute this every time you open R
```

It is important to create objects with explicit and short names, that are not already built-in functions in R. Ex: do not create a data named `data` or `summary` or `plot`. 

```{r import}
#tomato_data_2025_ILL_test_repeat<-read.csv("Rocio_tomato_modified.csv")
tomato <- read_xlsx('data/Rocio_tomato_modified.xlsx')
#tomato<-read.csv("data/Rocio_tomato_modified.csv")
```

Let's look at this new dataset:

```{r}
head(tomato)
summary(tomato)
```

## Formatting

Next, you will want to make sure that your variables are formatted adequately.
Here, we have variables that are numeric (allowing for mathematical operations) and some that are factors (with levels).
We will transform some in factors and take a look at the summary() to confirm the successful transformation.

```{r transform}
tomato$rep<-as.factor(tomato$rep)
tomato$Treatement<-as.factor(tomato$Treatement)
tomato$Infection<-as.factor(tomato$Infection)
summary(tomato)
```
We can observe quickly that a typo is now present in our data.
We could either correct it in the raw data file and reimport a new object, or correct it in R using the which() function and dimensionality operators.

```{r error}
which(tomato$Infection=="pathogn") #find the position of the error in the variable
tomato$Infection[10] #confirm that this is the error
tomato$Infection[10]<-"pathogen" #correct the error FOREVER by overwritting
tomato$Infection<-droplevels(tomato$Infection) #erase the empty factor level
summary(tomato) #check the summary again
```
## Visualizing the data

Histograms are a fundamental tool in R for exploratory data analysis (EDA). They help you:

**Visualize Distributions:** Quickly see the shape of your data (e.g., normal, skewed, bimodal).
**Spot Outliers & Gaps:** Identify unusual values or missing ranges in your dataset.
**Check Assumptions:** Assess if data meets statistical assumptions (e.g., normality for t-tests).
**Compare Groups:** Overlay histograms to compare different subsets (e.g., treatment vs. control).

In R, creating a histogram is simple (e.g., hist(data)), making it an essential first step in understanding your data before diving into analysis.

```{r hist}
hist(tomato$total.area)
```

```{r hist2}
par(mfrow=c(1,2))# Create a 1 x 2 plotting matrix
hist(tomato$total.area)
hist(tomato$chlorotic.area)
```

Scatterplot and boxplots are also very common graphs to be performed to explore the data. Here are a few examples:

```{r plot}
par(mfrow=c(2,2))# Create a 2 x 2 plotting matrix
# The next 4 plots created will be plotted next to each other
plot(tomato$total.area~tomato$Treatement)
plot(total.area~Treatement, data=tomato) #this is the same but the XY labels change
plot(total.area~Infection, data=tomato)
plot(chlorotic.area~Infection, data=tomato)
```

What happens with an interaction?

```{r interaction}
plot(total.area~Infection*Treatement, data=tomato)
```

Here is a scatterplot:

```{r scatterplot}
plot(total.area~chlorotic.area, data=tomato)
```

You can apply transformation in your plot formula:

```{r log}
par(mfrow=c(1,2))
plot(log(total.area)~Infection,data=tomato)
plot(I(total.area/100000)~Infection,data=tomato)
```

We can create new variables that combine Treatement and Infection as well as the ratio between both areas, then use it for plotting:

```{r combo}
tomato$combo<-as.factor(paste(tomato$Treatement,tomato$Infection,sep=""))
tomato$rate<-tomato$chlorotic.area/tomato$total.area
tomato$combo<-factor(tomato$combo,levels=levels(tomato$combo)[c(2,1,4,3)]) #here we reorder the levels
plot(rate~combo,data=tomato)
```

## Challenge 3

Create a new dataset from tomato but excluding rep #5.

Plot side-by-side the histogram of total.area with and without rep 5, as well as chlorotic.area with and without rep 5.

## Bonus

Create a dataset for each rep and plot the ratio between chlorotic.area/total.area for the for combined levels (combo).

## Hands on session 1

#### Create a new script for your project

#### Import your data

#### Explore it

#### Format your data (categorical data as factors)

#### Subset, compute group means and sd

#### Make a histogram of a numeric variable

#### Make a dot plot with two variables (at least one numeric)

#### Ask us how to do what you want to do !

# Plots with ggplot (Tuesday)

We work with a real dataset where tomato plants have been infected (or not) with a pathogen. We want to look at the proportion of leaf area affected by chlorosis and compare it between the infected and control groups.

## Objectives

1. Learn the tidyverse syntax

2. Plot the proportion of plant leaf covered by chlorosis with mean and standard devitions by replicate, in a way that will allow us to compare the two groups (infected and not infected) across the two treatments (mock of treated).


Two packages are required here, make sure you have installed them:

```{r}
# install.packages('tidyverse') # super-package that includes ggplot
# install.packages('openxlsx') # to parse excel files
```

## The tidyverse syntax

```{r load-packages, include=FALSE}
library(pacman)
p_load(tidyverse, magrittr, openxlsx, kableExtra)
```

```{r}

tomato <- read.xlsx('data/Rocio_tomato.xlsx')
str(tomato)

```

We introduce a modern syntax from the `tidyverse` superpackage, which uses the pipe operator `%>%`. You can print it by using the `shift+cmd+m` keyboard combination (or `shift+ctrl+m` on Windows).


### Understanding pipes %>% 

To show its usefuleness in making codes much easier to read, consider the following challenge: **Compute the mean proportion of chlorotic area of infected plants that were NOT treated, and format it to print as a percentage with 2 decimals.**.

A traditional base R code could look like this: 

```{r}
which_observations <- which(tomato$Infection == 'pathogen' & tomato$Treatement == 'mock')
inf_treat_chlor_area <- tomato$chlorotic.area[which_observations]
inf_treat_total_area <- tomato$total.area[which_observations]
prop_area <- 100*inf_treat_chlor_area/inf_treat_total_area

paste('Mean proportion:', round(mean(prop_area),2), "%")
```
Now compare the same thing with the tidyverse syntax. We will use 3 new functions, namely:

-`mutate()` is used to create new variables from existing ones

-`filter()` subsets our rows based on the values of variables

-`pull()` which returns a dataset column as a vector

```{r}
tomato %>% 
  filter(Infection == 'pathogen' & Treatement == 'mock') %>% 
  mutate(prop_area = 100*chlorotic.area/total.area) %>% 
  pull(prop_area) %>% 
  mean() %>% 
  round(2) %>% 
  paste("Mean proportion:", ., "%")
```

We can take advantage of that syntax to create more informative summaries of our data by using `group_by()`, a way to operate on subsets of our data without actually subsetting it; and passing the grouped dataframe to `summarise()`, which allows mathematical functions to be applied to every group. We then use `kable()` to format the output in a visually neat table.

```{r}
tomato_summary <- tomato %>% 
  mutate(prop_area = 100*chlorotic.area/total.area) %>% 
  group_by(Infection, Treatement) %>% 
  summarise(mean_prop = mean(prop_area), 
            sd_prop = sd(prop_area),
            n_obs = n(),
            .groups = 'drop')

tomato_summary
```

Doing this using base R syntax would have required many intermediate objects, whereas here we get the end product with a single pipe that does all we need.

We can also format this table if we wanted to export it. The kableExtra package is quite useful for this. We first round the values to the second decimal using `mutate()`, then change the column names and style the table.

```{r}
tomato_summary %>% 
  mutate(across(where(is.numeric), ~ round(., 2))) %>% 
  kable(col.names = c("Infection", "Treatment", "Mean % leaf chlorosis", "SD % leaf chlorosis", "N")) %>% 
  kable_styling("striped")
```


### Modifying variables

I have intentionally introduced a typographic error in the following dataset; factors will help us spot them quickly.To get there, let's convert characters as factors using `mutate()` :
```{r}

tomato_mdf <- read.xlsx('data/Rocio_tomato_modified.xlsx') %>% 
  mutate(Treatement = as.factor(Treatement),
         Infection = as.factor(Infection),
         rep = as.factor(rep)) # even if it's a number, it represents categorical information
summary(tomato_mdf)
```

Consider the following things we might want to do as well:

1.  The `Treatement` column name has an typo (an extra `e`) in it, it should be `Treatment`
2.  There is a typo in one of the Infection values we want to correct
3.  We want to compute the proportion of the total area that is chlorotic

```{r}

tomato_mdf <- tomato_mdf %>% 
  rename(Treatment = Treatement) %>% 
  mutate(
    prop_area = 100*chlorotic.area/total.area,
    Infection = as.factor(
      case_when(
        Infection == "pathogn" ~ "pathogen",
        TRUE ~ Infection
        ))
    )

summary(tomato_mdf)
dim(tomato_mdf)
```

### Basic string manipulation

We could go further and modify text values in our dataset. For example, we might want to capitalise the first letter of `pathogen`. We could use a case_when(), but if we want to capitalise all values in our variables, it becomes cumbersome. Instead, we can use one of the many string manipulation verbs in the `tidyverse` package.

```{r}
tomato_mdf %<>% 
  mutate(
    Infection = str_to_title(Infection),
    Treatment = str_to_upper(Treatment)
  )
head(tomato_mdf)
```

Here I introduced the assignment pipe operator `%<>%`. It is equivalent to doing `tomato_mdf <- tomato_mdf %>% ...`. To use that pipe, the `magrittr` package must be explicitly loaded (we did it earlier).

### Pivoting tables

### 

```{r }

```

# Stats in R (Wednesday)
