---
title: "Session 4: Making better plots"
author: "Isabelle Laforest-Lapointe"
date: "2025-05-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup: load packages

```{r packages,message=F,warning=F}
library(readxl)
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
library(gapminder) # For demo dataset
```

## Lesson Objectives

By the end of this session, you will:

Understand ggplot2's theme system
Apply built-in and custom themes
Combine multiple plots with patchwork
Create publication-quality figure layouts

## GGPLOT2 Themes

You can find all the possible ggplot2 themes here: https://ggplot2.tidyverse.org/reference/ggtheme.html 

```{r themes}
data <- gapminder %>% filter(year == 2007)

p <- data %>%
  ggplot(aes(gdpPercap, lifeExp)) +
  geom_point(aes(size = pop, color = continent)) +
  scale_x_log10() +
  labs(title = "GDP vs Life Expectancy",
       x = "GDP per capita",
       y = "Life Expectancy")

# Demonstrate theme options
p + theme_dark()
p + theme_gray()     # Default
p + theme_bw()       # Black-and-white
p + theme_minimal()  # Clean minimal
p + theme_classic()  # Traditional
p + theme_void()     # Empty canvas
```

## Customizing your theme

You can modify greatly all aspects of your figures.
There are numerous resources online to browse colours in R but here is one that I like to use: https://colorbrewer2.org/#type=sequential&scheme=Reds&n=3.

```{r custom}
p + theme(
  plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
  panel.background = element_rect(fill = "aliceblue"),
  legend.position = "bottom",
  axis.text = element_text(color = "darkblue")
)
```
## Facetwraps

Now we will do this figure (boxplot) step by step to learn how to incrementally build a ggplot2 figure.

Let's first reload the tomato (clean) data and relabel the factors to practice our newly learned syntax.

```{r tomato}
#Load the data
tomato<-read_xlsx("Rocio_tomato_modified.xlsx",sheet=2)
#Prepare the relabel function for both variables for clean plotting
relabel.Treatement = as_labeller(c(mock = "Mock", 
                   treated = "Treated"))
relabel.Infection = as_labeller(c(pathogen = "Pathogen", 
                   water = "Water"))
#Transform our dataset with the %<>%
tomato %<>%
  rowwise() %>% 
  mutate(Treatement = factor(relabel.Treatement(Treatement))) %>%
  mutate(Infection = factor(relabel.Infection(Infection)))
```

Okay, let's start to build our plot!

```{r 1}
tomato %>%
  ggplot(aes(Infection,chlorotic.area/total.area,shape=Infection))+
  geom_boxplot(aes(fill=Infection))
```
Let's add the facet!

```{r 2}
(first_plot <- tomato %>%
  ggplot(aes(Infection,chlorotic.area/total.area,shape=Infection))+
  geom_boxplot(aes(fill=Infection))+
  facet_wrap(~Treatement))
```
Great, now we can change our theme to the dark side!

```{r 3}
tomato %>%
  ggplot(aes(Infection,chlorotic.area/total.area,shape=Infection))+
  geom_boxplot(aes(fill=Infection))+
  facet_wrap(~Treatement)+
  theme_dark()
```

Let's add a jitter to see better each observation.

```{r 4}
tomato %>%
  ggplot(aes(Infection,chlorotic.area/total.area,shape=Infection))+
  geom_boxplot(aes(fill=Infection))+
  facet_wrap(~Treatement)+
  theme_dark()+
  geom_jitter(width = 0.1,aes(fill=Treatement))
```

Let's change the colours and the shape of our point to be fillable.

```{r 5}
tomato %>%
  ggplot(aes(Infection,chlorotic.area/total.area,shape=Infection))+
  geom_boxplot(aes(fill=Infection))+
  facet_wrap(~Treatement)+
  theme_dark()+
  geom_jitter(width = 0.1,aes(fill=Treatement))+
  scale_fill_manual(values = c("#69B3A2","#de2d26","white","white"))+
  scale_shape_manual(values = c(21,21))+
  ylab(label="Leaf chlorotic area (%)")+
  guides(fill=F,shape=F)
```

Perfect, now on to changing the Y label to something more publication ready.
We can also remove the legend.

```{r 6}
(final_plot <- tomato %>%
  ggplot(aes(Infection,chlorotic.area/total.area,shape=Infection))+
  geom_boxplot(aes(fill=Infection))+
  facet_wrap(~Treatement)+
  theme_dark()+
  geom_jitter(width = 0.1,aes(fill=Treatement))+
  scale_fill_manual(values = c("#69B3A2","#de2d26","white","white"))+
  scale_shape_manual(values = c(21,21))+
  ylab(label="Leaf chlorotic area (%)")+
  guides(fill=F,shape=F))
```

## Challenge 7

Create your own version of this graph, but changing the theme, the colours, the shape of the points, the X and Y labels, and adding a main title.

## Patchwork

Patchwork is an amazing package that allows you to create perfect multi-plot figures for your publications.

Here is an overview of the possibilities:

```{r patchwork}
p1 <- ggplot(mpg, aes(displ, hwy)) + geom_point()
p2 <- ggplot(mpg, aes(class)) + geom_bar()
p3 <- ggplot(mpg, aes(cty)) + geom_histogram()

# Horizontal layout
p1 + p2

# Vertical layout
p1 / p2

# Grid layout
(p1 | p2) / p3
```

Here are some additional options:

```{r annotate,message=F,warning=F}
# Add title to composition
(p1 + p2) + plot_annotation(title = "My Dashboard")

# Control spacing and guides
(p1 + p2 + p3) + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```

## Challenge 8

Using the boxplot figures that we created from the tomato dataset, combine three different plots (the first one and the final one that we created together + the one you modified)

Combine them into a 1x3 horizontal layout
Add a shared title "Walk through a tomato plotting contest"
Label subplots as A, B, C