---
title: "Session 2: Subsets and Plots"
author: "Isabelle Laforest-Lapointe"
date: "2025-05-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Importing data

First, you want to import your data.
It is important to create objects with explicit and short names, that are not already built-in functions in R.
Ex: do not create a data named data or summary or plot.

```{r import}
#tomato_data_2025_ILL_test_repeat<-read.csv("Rocio_tomato_modified.csv")
tomato<-read.csv("Rocio_tomato_modified.csv")
```
## Exploring data

The function summary() is very useful to explore what is in your data and also to confirm that your data was imported well.

```{r summary}
summary(tomato)
```
The head() and View() functions are also good ways to take a peek at your data.

```{r head}
head(tomato)
```
## Formatting

Next, you will want to make sure that your variables are formatted adequately.
Here, we have variables that are numeric (allowing for mathematical operations) and some that are factors (with levels).
We will transform some in factors and take a look at the summary() to confirm the successful transformation.

```{r transform}
tomato$rep<-as.factor(tomato$rep)
tomato$Treatement<-as.factor(tomato$Treatement)
tomato$Infection<-as.factor(tomato$Infection)
summary(tomato)
```
We can observe quickly that a typo is now present in our data.
We could either correct it in the raw data file and reimport a new object, or correct it in R using the which() function and dimensionality operators.

```{r error}
which(tomato$Infection=="pathogn") #find the position of the error in the variable
tomato$Infection[10] #confirm that this is the error
tomato$Infection[10]<-"pathogen" #correct the error FOREVER by overwritting
tomato$Infection<-droplevels(tomato$Infection) #erase the empty factor level
summary(tomato) #check the summary again
```
## Visualizing the data

Histograms are a fundamental tool in R for exploratory data analysis (EDA). They help you:

**Visualize Distributions:** Quickly see the shape of your data (e.g., normal, skewed, bimodal).
**Spot Outliers & Gaps:** Identify unusual values or missing ranges in your dataset.
**Check Assumptions:** Assess if data meets statistical assumptions (e.g., normality for t-tests).
**Compare Groups:** Overlay histograms to compare different subsets (e.g., treatment vs. control).

In R, creating a histogram is simple (e.g., hist(data)), making it an essential first step in understanding your data before diving into analysis.

```{r hist}
hist(tomato$total.area)
```

```{r hist2}
par(mfrow=c(1,2))# Create a 1 x 2 plotting matrix
hist(tomato$total.area)
hist(tomato$chlorotic.area)
```

Scatterplot and boxplots are also very common graphs to be performed to explore the data.
Here are a few exemples:

```{r plot}
par(mfrow=c(2,2))# Create a 2 x 2 plotting matrix
# The next 4 plots created will be plotted next to each other
plot(tomato$total.area~tomato$Treatement)
plot(total.area~Treatement, data=tomato) #this is the same but the XY labels change
plot(total.area~Infection, data=tomato)
plot(chlorotic.area~Infection, data=tomato)
```

What happens with an interaction?

```{r interaction}
plot(total.area~Infection*Treatement, data=tomato)
```

Here is a scatterplot:

```{r scatterplot}
plot(total.area~chlorotic.area, data=tomato)
```

You can apply transformation in your plot formula:

```{r log}
par(mfrow=c(1,2))
plot(log(total.area)~Infection,data=tomato)
plot(I(total.area/100000)~Infection,data=tomato)
```

We can create new variables that combine Treatement and Infection as well as the ratio between both areas, then use it for plotting:

```{r combo}
tomato$combo<-as.factor(paste(tomato$Treatement,tomato$Infection,sep=""))
tomato$rate<-tomato$chlorotic.area/tomato$total.area
tomato$combo<-factor(tomato$combo,levels=levels(tomato$combo)[c(2,1,4,3)]) #here we reorder the levels
plot(rate~combo,data=tomato)
```

## Challenge 3

Create a new dataset from tomato but excluding rep #5.

Plot side-by-side the histogram of total.area with and without rep 5, as well as chlorotic.area with and without rep 5.

## Bonus

Create a dataset for each rep and plot the ratio between chlorotic.area/total.area for the for combined levels (combo).