---
title: "Session 1 - Introduction"
output: html_document
---

## RStudio environement

When you open RStudio, you will have four panels:  

**Top left:** Script  
Document where you will write the command lines:This document can be saved for further modifications.  

**Bottom left:** Console  
Where the output of the command line will appear. Use it to test stuff that you do not want to save in the script.

**Top right:** Environment  
Place where objects (dataframes, values, vectors) are stored. Click on any object to visualise it.

**Bottom right:** Files, Plots, Help  
This is a files and plots viewer as well as the help center.

## Basic operations 
Use R script as a calculator and visualize the output in the console
```{r}
# Everything written after a hashtag will be considered as a comment (Not a command line)
2+4
```
Store the output in an object
```{r}
res <- 2+4
res
```
Multiply the result
```{r}
2*res
```

## Data types
Character  
Numeric  
Integer  
Factor  
Logical  
Vector  
Data frame  
Matrix
```{r}
# Create a vector of numeric values
vec1 <- c(1,3,5)
# Create a vector of characters values
vec2 <- c("one","three","five")
# Create a vector of logical values
vec3 <- c(TRUE, TRUE, FALSE)
# Bind these vectors into a dataframe
df1 <- data.frame(vec1, vec2, vec3)
```
```{r}
class(vec1) # Use function class( ) to get the class of an object
class(vec2)
vec2 <- as.factor(vec2) # Change the class of vec2 to factor
class(vec2)
class(df1)
```

## Common symbols
TRUE/FALSE  
**<-** assign an object  
**=** assign an object and equal  
**==** equal (when sub-setting)  
**<** smaller than  
**>** greater than  
**<=** smaller or equal to  
**>=** greater or equal to  
**!=** is not equal  
```{r}
5 > 6
5 <= 6
5 != 4+1
```

## Save an object
```{r}
# Save object
saveRDS(df1, "df1.RDS")
# Read/open object
readRDS("df1.RDS")
```






