---
title: "Day3_AM"
output: html_document
date: "2025-05-01"
---

## Packages
Run the `install.packages()` if packages are not installed yet.
```{r results='hide', message=FALSE}
#install.packages("ggpubr")
library(ggpubr)
#install.packages("ggplot2")
library(ggplot2)
#install.packages("FSA")
library(FSA)
#install.packages("dunn.test")
library(dunn.test)
#install.packages("dplyr")
library(dplyr)
#install.packages("rstatix")
library(rstatix)
#install.packages("car")
library(car)
#install.packages("readxl")
library(readxl)
```

## Compare the mean across more than two groups

Let's use the Moss data set and plot the Nitrogen fixation (ARA_C2H4) across sampling months in 2017 independently for each moss species.

```{r}
# Load data
Moss <- read_excel("data/Marie_Mosses.xlsx")
# Transform some variables as factors
Moss$Year <- as.factor(Moss$Year)
Moss$Months <- factor(Moss$Months, levels = c("June","September","October"))
# Subset to keep only data from 2017
Moss2017 <- Moss[Moss$Year=="2017",]
# Subset to keep only data from one moss specie
Moss2017_PCC <- Moss2017[Moss2017$Species == "Ptilium crista-castrensis",]
```

```{r}
# Plot it
ggplot(data = Moss2017_PCC, aes(x = Months, y = ARA_C2H4, fill = Months))+
  geom_boxplot()+
  theme_bw()+
  scale_fill_manual(values = c("green","green4", "darkgreen"))
```

We want to evaluate if there is a significant difference in the nitrogen fixation across time. Therefore, we want to compare the mean of 3 groups (June, September, October).

## ANOVA (parametric)

**Conditions to apply anova (parametric)** 

1. Independent observation  
2. Normality (`shapiro.test()`) 
```{r}
shapiro.test(Moss2017_PCC$ARA_C2H4[Moss2017_PCC$Months == "June"])
# p>0.05: data is normal
shapiro.test(Moss2017_PCC$ARA_C2H4[Moss2017_PCC$Months == "September"])
# p>0.05: data is normal
shapiro.test(Moss2017_PCC$ARA_C2H4[Moss2017_PCC$Months == "October"])
# p>0.05: data is normal
```
3. Homogeneous variance (`leveneTest()`) 
```{r}
leveneTest(Moss2017_PCC$ARA_C2H4 ~ Moss2017_PCC$Months)
# p>0.05: variances are not significantly different
```
#### Perform Anova
```{r}
# Using anova()
res <- anova(lm(Moss2017_PCC$ARA_C2H4 ~ Moss2017_PCC$Months))
res # No difference between groups (Months)

# Using aov()
res2 <- aov(Moss2017_PCC$ARA_C2H4 ~ Moss2017_PCC$Months)
summary(res2) # No difference between groups (Months)
```
According to the anova, there is no difference between sampling months.  

#### Tukey test (parametric post-hoc)
If the anova had revealed a difference, we would have performed a Tukey test (parametric post-hoc) to know between which group there is a difference.
```{r}
# Difference between which groups?
TukeyHSD(res2) # No difference between groups (Months)
```
## Kruskal-Wallis (non-parametric)

Let's use the flowers data set and plot the alpha-diversity for each site (code)
```{r}
# Import dataset
flowers <- read_excel("data/meta.flower.2022.xlsx")
# Transform the variable code as factor
flowers$code <- as.factor(flowers$code)

# Plot alpha-diversity for each site (code)
(plot.flower <- ggplot(flowers, aes(code, alphadiv, fill = code))+
  geom_boxplot()+
    theme_bw())
```

Test normality
```{r}
shapiro.test(flowers$alphadiv[flowers$code == "A"])
# p>0.05: data is normal
shapiro.test(flowers$alphadiv[flowers$code == "B1"])
# p>0.05: data is NOT normal
shapiro.test(flowers$alphadiv[flowers$code == "C"])
# p>0.05: data is NOT normal
```
#### Perform Kruskal test
```{r}
kruskal_test(flowers, alphadiv~code)
# there is a significant difference between some groups
```

#### Dunn test (non-parametric post-hoc)
```{r}
dunn_test(flowers, alphadiv~code)
# Differences are between groups A-C and B1-C
# No difference between groups A-B1
```

Let's plot the statistical result on the graph
```{r}
# Store dunn test result in an object
dunn.res <- dunn_test(flowers, alphadiv~code)
# Manually extract p-values from dunn test results
p_values <- dunn.res$p.adj.signif[c(1, 2, 3)]  
# Find x and y positions for the p-value
stat.flower <- add_xy_position(dunn.res, x = "code")
# List of comparison
comparisons <- list(c("A", "B1"), c("A", "C"), c("B1", "C"))
# Add on the graph
plot.flower+
  geom_signif(comparisons = comparisons,
              annotations = p_values,
              y_position = stat.flower$y.position) 
```

To conclude, if we want two compare the mean between more than two groups we first need to know if we use a parametric (Anova) or a non-parametric (Kruskal-Wallis) test. So we first test the normality of all group using `shapiro.test()` (p<0.05: Not normal; p>0.05 Normal) and the homogeneity of the variance using `leveneTest()`.  

Depending on the results, we will perform an Anova or a Kruskal test. If the result of the Anova is significant, we then perform a post-hoc Tukey test. If the the result of the Kruskal test is dignificant, we then perform a post-hoc Dunn test.


## Correlation between two variables

### Pearson & Spearman

Pearson = parametric  
Spearman = non-parametric

Test normality
```{r}
shapiro.test(Moss2017$Fe_ppm)
# p<0.05: Not normal
shapiro.test(Moss2017$P_ppm)
# p>0.05 Normal
shapiro.test(Moss2017$CN_Ratio)
# p>0.05 Normal
shapiro.test(Moss2017$`N_mg.g-1`)
# p>0.05 Normal
```
Perform correlation test
```{r}
# Pearson (parametric)
plot(Moss2017$P_ppm,Moss2017$CN_Ratio)
cor.test(Moss2017$P_ppm, Moss2017$CN_Ratio, method = "pearson")
# p<0.05: Reject H0 -> there is a correlation (r = -0.4)

# Spearman (non-parametric)
plot(Moss2017$Fe_ppm, Moss2017$P_ppm)
cor.test(Moss2017$Fe_ppm, Moss2017$P_ppm, method = "spearman")
# p<0.05: Reject H0 -> there is a correlation (r = -0.5)

# Pearson (parametric)
plot(Moss2017$`N_mg.g-1`, Moss2017$CN_Ratio)
cor.test(Moss2017$`N_mg.g-1`, Moss2017$CN_Ratio, method = "pearson")
# p<0.05: Reject H0 -> there is a correlation (r = -0.98)
```

Add correlation coefficient and p-value to the graph
```{r}
ggplot(Moss2017, aes(`N_mg.g-1`,CN_Ratio, add = "reg.line"))+
  geom_point()+
  theme_bw()+
  geom_smooth(method = "lm")+
  stat_cor(method = "pearson", label.x = 9, label.y = 63)
```

# Challenge #10  

Test if there is a correlation between two other variables from the moss data set  
- Test normality of both variables  
- Perform appropriate correlation test  
- Plot it and add correlation coefficient and p-value

