---
title: 'Session 6: Linear models +'
author: "Isabelle Laforest-Lapointe"
date: "2025-05-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Objectives

By the end of this session, you will:

**1. Understand When to Use Mixed/NL Models**

Identify scenarios requiring random effects (LMMs) or nonlinear relationships (e.g., growth curves, repeated measures).

**2. Build & Interpret LMMs in R**

Fit models with lme4/nlme (e.g., lmer(y ~ x + (1|ID))), interpret random vs. fixed effects, and evaluate assumptions (residual plots, singularity).

**3. Model Nonlinear Trends**

Implement flexible approaches: polynomials (poly(x,2)), splines (ns()), or GAMs (mgcv::gam) for curved relationships.

**4. Visualize Results Effectively**

Create diagnostic plots (QQ, residuals) and publication-ready graphs with ggplot2/patchwork, including random effects predictions and confidence bands.

## Load packages

Here are all the packages that you need to load for this session, you can either do it one by one, the *long* way:

```{r long_packages, message=F, warning=F}
library(tidyverse)
library(patchwork) # For beautiful plots
library(lme4)    # For mixed models
library(lmerTest) # To add p-values
library(emmeans) # For post-hoc analyses
```

Or you can use the function **p_load** in the package **pacman** to load all packages that you need at the same time. p_load allows to install and load at the same time, which is **great**!

```{r short_packages, message=FALSE, warning=F}
library(pacman) # A magician package to load packages
pacman::p_load(c(tidyverse,patchwork,lme4,lmerTest,emmeans))
```
## Load roo data

```{r data, message=FALSE, warning=FALSE}
roo <- read_csv("roo.csv")
dim(roo)
```

Let's take a quick peek before starting:

```{r summary}
summary(roo)
```

You can also use the function *View()* to see the whole dataset.

Now, let's modify it a bit for the sake of simplicity. We will remove all individuals that have Ad in the Age variable. Here is a code using base R to do it but you can also use dplyr/tidyr syntax to do so:

```{r Age}
roo<-roo[-c(which(roo$Age=="Ad")),];dim(roo)
```

We will also filter out any roo that does not have 5 measurements (recaptures):

```{r 5}
roo_filtered <- roo %>%
  group_by(ID) %>%          # Group by the ID column
  filter(n() >= 5) %>%      # Keep groups with 5+ rows
  ungroup()
dim(roo_filtered)
```

So, we will explore the relationship between mass, age, and sex.
Before continuing to the models, we need to make sure our variables are good to go.
Age needs to be numeric
Sex and ID need to be factors

```{r tranformations}
roo_filtered$Age <- as.numeric(roo_filtered$Age)
roo_filtered$Sex <- as.factor(roo_filtered$Sex)
roo_filtered$ID <- as.factor(roo_filtered$ID)
```

Let's confirm everything looks good:

```{r confirm}
summary(roo_filtered)
```

Excellent, now on to our Y variable:

```{r mass}
summary(roo_filtered$Mass)
hist(roo_filtered$Mass)
```

We are ready to go!

## Fit a Linear Model

**Formula Breakdown:**
Mass ~ Age * Sex: Tests main effects of Age and Sex, plus their interaction

```{r lm_model}
model_lm <- lm(Mass ~ Age * Sex, data = roo_filtered)
summary(model_lm)
```

## Fit a Linear Mixed Model

**Formula Breakdown:**
Mass ~ Age * Sex: Tests main effects of Age and Sex, plus their interaction
(1 | ID): Random intercept for each kangaroo (accounts for repeated measures)

```{r lmr_model}
model_lmr <- lmer(Mass ~ Age * Sex + (1 | ID), data = roo_filtered)
summary(model_lmr)
```

The model summary shows:

Fixed effects (Age, Sex, Age:Sex) with estimates and p-values

Random effects variance (how much IDs deviate from the mean)

Model fit statistics (AIC, BIC)

## Residual diagnostics

Let's check the residuals diagnostic of the model to confirm they are good

```{r diagnostic}
plot(model_lmr)               # Residuals vs. fitted
qqnorm(resid(model_lmr))      # QQ plot
hist(resid(model_lmr))        # Residual distribution
```

## Check for singularity:

Did you over-fit your model:

```{r singular}
isSingular(model_lmr)  # Should return FALSE
```

## Post-hoc Analysis (If Interaction is Significant)

Here we will check simple slopes analysis.

```{r posthoc}
emtrends(model_lmr, ~ Sex, var = "Age")  # Age effect per sex level
```

And perform pairwise comparisons:

```{r pair}
emm <- emmeans(model_lmr, ~ Age * Sex)
pairs(emm, adjust = "tukey")
```

## Alternative Model Specifications

You could also want random slopes:

```{r slopes}
model_rs <- lmer(Mass ~ Age * Sex + (Age | ID), data = roo_filtered)
summary(model_rs)
```

## Plotting the linear model

```{r plot_lm, warning=F,message=F}
A<-ggplot(roo_filtered, aes(Age, Mass, color = Sex)) + 
  geom_point() + scale_color_manual(values = c("#404080", "#69B3A2"))+
  theme_minimal()+
  geom_smooth(method = "lm")+
  labs(title = "Linear fit: Mass vs. Age by Sex")+
  theme(
    legend.position = c(0.95, 0.75),            # Bottom-right (0-1 coordinates)
    legend.justification = c(1, 0),              # Anchor point
    legend.box.background = element_rect(color = "black", size = 0.5),  # Border
    legend.box.margin = margin(5, 5, 5, 5)       # Padding
  );A
```

Oh, now we realize that the relationship between Age and Mass appears non-linear in the ggplot, we could thus adapt our analysis to use a non-linear model instead to fit better the data:

## Basic GAM (Non-linear Fixed Effects)

```{r mgcv}
library(mgcv)
```

## Fit the GAM

Fit GAM with smooth term for Age, fixed effect for Sex, and their interaction

```{r GAM}
gam_model <- gam(
  Mass ~ s(Age) + Sex + s(Age, by = Sex),  # s() = smooth term
  data = roo_filtered,
  method = "REML"  # Restricted Maximum Likelihood (default)
)
summary(gam_model)
```

In this code:

s(Age): Smooth (non-linear) effect of Age.

by = Sex: Allows the smooth effect of Age to vary by Sex.

## GAM with Random Effects (GAMM)

To account for repeated measures (ID as a random effect):

# Using mgcv (simpler syntax)
```{r model2}
gamm_model <- gam(
  Mass ~ s(Age) + Sex + s(Age, by = Sex) + s(ID, bs = "re"),  # bs="re" = random effect
  data = roo_filtered,
  method = "REML"
)
summary(gamm_model)
```

## Model Diagnostics

# Check smooth terms and residuals
```{r gam_model}
plot(gamm_model, pages = 1, shade = TRUE)  # Visualize smooths
gam.check(gamm_model)                      # Diagnostic plots (QQ, residuals vs. fitted)
```

## Compare to Linear Model

# Fit linear model for comparison
```{r comparison}
lm_model <- lm(Mass ~ Age * Sex, data = roo_filtered)
```

# Compare AIC (lower = better)
```{r AIC}
AIC(gam_model, lm_model)
```

# Plot it

```{r plot_nlm,warning=F,message=F}
B<-ggplot(roo_filtered, aes(Age, Mass, color = Sex)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "gam", formula = y ~ s(x), se = TRUE) +  # GAM smoother+
  scale_color_manual(values = c("#404080", "#69B3A2"))+
  theme_minimal()+
  labs(title = "Non-linear fit: Mass vs. Age by Sex")+
  theme(
    legend.position = c(0.85, 0.95),
    legend.direction = "horizontal",                    # Stack items horizontally
    legend.title = element_text(size = 10, face = "bold"),
    legend.text = element_text(size = 8),
    legend.key = element_blank()                       # Remove key background
  );B
```

## Patchwork
And we can wrap up this session by merging the two plots with patchwork:

```{r patchwork, message=FALSE,warning=F}
A+B+plot_annotation(tag_levels = 'A') &
  theme(plot.tag = element_text(size = 14, face = "bold"))
```

## Challenge 12

Create a subset dataset with the roos Age < 6 and fit a linear mixed model to the data with the same formula that we used above.
Look at the model summary and perform the diagnostics.


