---
title: 'Session 5: A primer for statistics'
author: "Isabelle Laforest-Lapointe"
date: "2025-05-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set your working directory

The functions setwd() and getwd() are your friends.

Navigate in the Files panel with the blue wheel make sure that you see your working directory.

## Load the packages you need

```{r packages,message=F,warning=F}
library(tidyverse)
library(readxl)
```

## Look at your data

To start, we will load the tomato data and look at it.

```{r tomato data}
tomato<-read_xlsx("Rocio_tomato_modified.xlsx",sheet=2)
tomato$Infection<-as.factor(tomato$Infection)
tomato$Infection[which(tomato$Infection=="pathogn")]<-"pathogen"
tomato$Infection<-droplevels(tomato$Infection)
tomato <- tomato %>%
  mutate(ratio = chlorotic.area / total.area)

pathogen_only <- tomato %>%  
  filter(Infection == "pathogen")
```

What should we do first?

```{r hist}
hist(pathogen_only$ratio)
hist(pathogen_only$total.area)
hist(pathogen_only$chlorotic.area)
```
## Normality

Then we could check for normality:

```{r normal}
shapiro.test(pathogen_only$ratio)
```
OH MY GOD a p-value, what does this mean?
Because it is <0.05, we can reject our H0 that the data is distributed normally.
Thus, our data does not appear to follow a normal distribution.

Here is a different way to perform normality tests on all the variables you want:

```{r mnormal}
variables_to_test <- c("chlorotic.area", "total.area", "ratio")

# Loop over variables and apply Shapiro-Wilk test
shapiro_results <- tomato %>%
  select(all_of(variables_to_test)) %>%  # Select only the target variables
  map(~ shapiro.test(.x)) %>%            # Apply Shapiro-Wilk test to each column
  map_dfr(~ broom::tidy(.x), .id = "variable")  # Convert results to a tidy data frame

# Print results
print(shapiro_results)
```

## Homoskedasticity

The second assumption is equal variance in our data, let's test it:

```{r var}
var.test(ratio~Treatement,data=pathogen_only)
```

Okay, so let's imagine our data is normal, what is the next step, look at our data again:

```{r plot}
boxplot(ratio~Treatement,data=pathogen_only) # a quick plot
pathogen_only %>% ggplot(aes(Treatement, ratio, fill=Treatement, shape=Treatement))+
  geom_boxplot()+geom_jitter(width=0.1,aes(fill="white"),size=3)+theme_classic()+
  scale_shape_manual(values=c(21,21))+
  scale_fill_manual(values = c("#69B3A2","#de2d26","white"))+
  labs(title="Ratio of chlorotic and total areas impacted by treatment",x="Treatment",y="Ratio")+
  guides(fill="none",color="none",shape="none")
```

In this graph, our H0 is that there is no difference in the mean of the Y variable between the two groups.
Our H1 could be that there is a difference (less or more), or we could already have hypothesized that the treated group will have a lower ratio because of literature or our previous knowledge.
Thus, we could expect significant results, no?
So let's run a t-test:

```{r ttest}
t.test(ratio~Treatement,data=pathogen_only) # R defaults to the unequal variance test Welch Two Sample t-test
t.test(ratio~Treatement,data=pathogen_only, var.equal=T) # default is two-sided# default is two-sided
t.test(ratio~Treatement,data=pathogen_only,alternative = "less",var.equal=T)
t.test(ratio~Treatement,data=pathogen_only,alternative = "greater",var.equal=T)
```

Let's now run the non-parametric equivalent, the Mann-Whitney-Wilcoxon Test:

```{r mann}
wilcox.test(ratio~Treatement,data=pathogen_only)
```

Still significant, we can now reject our H0.
It seems like the means of the two groups are not equal.
Did we demonstrate H1? No! But we cannot accept H0.

## Challenge 9

Create a water_only dataset
Look at the histogram of the ratio variable
Check for the normality and homoskedasticity of the ratio variable
Perform a t.test and a Mann-Whitney-Wilcoxon
