---
title: "Session 4: Making better plots"
author: "Isabelle Laforest-Lapointe"
date: "2025-05-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Setup: load packages

```{r packages, include=FALSE}
library(ggplot2)
library(patchwork)
library(tidyr)
library(dplyr)
library(gapminder) # For demo dataset
```

## Lesson Objectives

By the end of this session, you will:

Understand ggplot2's theme system
Apply built-in and custom themes
Combine multiple plots with patchwork
Create publication-quality figure layouts

## GGPLOT2 Themes

You can find all the possible ggplot2 themes here: https://ggplot2.tidyverse.org/reference/ggtheme.html 

```{r cars}
data <- gapminder %>% filter(year == 2007)

p <- ggplot(data, aes(gdpPercap, lifeExp)) +
  geom_point(aes(size = pop, color = continent)) +
  scale_x_log10() +
  labs(title = "GDP vs Life Expectancy",
       x = "GDP per capita",
       y = "Life Expectancy")

# Demonstrate theme options
par(mfrow=c(3,2))
p + theme_dark()
p + theme_gray()     # Default
p + theme_bw()       # Black-and-white
p + theme_minimal()  # Clean minimal
p + theme_classic()  # Traditional
p + theme_void()     # Empty canvas
```

## Customizing your theme

You can modify greatly all aspects of your figures.

```{r custom}
p + theme(
  plot.title = element_text(size = 16, face = "bold", hjust = 0.5),
  panel.background = element_rect(fill = "aliceblue"),
  legend.position = "bottom",
  axis.text = element_text(color = "darkblue")
)
```

## Challenge 7

Create a scatterplot of your choice using the mpg dataset and:

Apply theme_light() as base
Customize:
Make title center-aligned and red
Change panel grid to dotted lines
Move legend inside plot (position at c(0.8, 0.8))

## Patchwork

Patchwork is an amazing package that allows you to create perfect multi-plot figures for your publications.

Here is an overview of the possibilities:

```{r layout}
p1 <- ggplot(mpg, aes(displ, hwy)) + geom_point()
p2 <- ggplot(mpg, aes(class)) + geom_bar()
p3 <- ggplot(mpg, aes(cty)) + geom_histogram()

# Horizontal layout
p1 + p2

# Vertical layout
p1 / p2

# Grid layout
(p1 | p2) / p3
```

Here are some additional options:

```{r layout}
# Add title to composition
(p1 + p2) + plot_annotation(title = "My Dashboard")

# Control spacing and guides
(p1 + p2 + p3) + 
  plot_layout(guides = "collect") +
  plot_annotation(tag_levels = "A")
```

## Challenge 8

Using the diamonds dataset:

Create three different plots (any types)
Combine them into a 1x3 horizontal layout
Add a shared title "Diamond Analysis"
Label subplots as A, B, C